---
title: "HRI AGRRA Belize — EDA"
subtitle: "Coral reef response variables & oceanographic predictors"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    fig-width: 9
    fig-height: 5.5
    code-fold: true
    code-summary: "Code"
    embed-resources: true
    smaller: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup}
#| echo: false
library(tidyverse)
library(here)
library(corrplot)

fig_dir <- here("results", "figures")

theme_hri <- function() {
  theme_minimal(base_size = 16) +
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 20),
      plot.subtitle = element_text(hjust = 0.5, size = 14, color = "gray50"),
      axis.title.y  = element_text(face = "bold"),
      axis.text.x   = element_text(angle = 45, vjust = 0.5),
      panel.grid.major = element_line(color = "gray85", linetype = "dashed"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 12)
    )
}

full  <- read.csv(here("results", "tables", "ResponseVariables_full.csv"))
input <- read.csv(here("results", "tables", "ResponseVariables_input.csv"))
full$YEAR <- factor(full$YEAR)
```

## Data Overview

:::: {.columns}

::: {.column width="50%"}
**Full dataset**

- `r nrow(full)` observations (site x year)
- `r ncol(full)` variables
- Years: 2018, 2021, 2023

**Site-averaged**

- `r nrow(input)` unique reef sites
- 4 response variables
:::

::: {.column width="50%"}
```{r}
#| echo: false
full %>%
  group_by(YEAR) %>%
  summarise(
    `Total` = n(),
    `Benthic` = sum(!is.na(tCORALavg)),
    `Fish` = sum(!is.na(tCommBiomass)),
    .groups = "drop"
  ) %>%
  knitr::kable()
```
:::

::::

## Missing Data

```{r}
#| echo: false
na_pct <- colMeans(is.na(full)) * 100
na_df  <- tibble(variable = names(na_pct), pct = na_pct) %>%
  filter(pct > 0) %>% arrange(desc(pct))

ggplot(na_df, aes(x = reorder(variable, pct), y = pct)) +
  geom_col(fill = "#2C3E50") + coord_flip() +
  labs(x = NULL, y = "% Missing", title = "Missingness by Variable") +
  theme_hri()
```

::: {.notes}
Benthic and fish surveys don't always overlap at the same sites.
Reef Type is only available for 2023 data.
:::

## Benthic Cover — Distributions

```{r}
benthic_long <- full %>%
  filter(!is.na(tCORALavg)) %>%
  pivot_longer(c(tCORALavg, tFMAavg), names_to = "var", values_to = "val") %>%
  mutate(var = recode(var, tCORALavg = "Coral", tFMAavg = "Macroalgae"))

ggplot(benthic_long, aes(x = YEAR, y = val, fill = var)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_manual(values = c("#1b9e77", "#d95f02")) +
  labs(x = NULL, y = "Cover (%)", fill = NULL,
       title = "Benthic Cover by Year") +
  theme_hri()
```

## Benthic Cover — Trends

```{r}
avg_b <- full %>% filter(!is.na(tCORALavg)) %>%
  group_by(YEAR) %>%
  summarise(coral = mean(tCORALavg), algae = mean(tFMAavg), .groups = "drop")

ggplot(avg_b, aes(x = YEAR, group = 1)) +
  geom_line(aes(y = coral, color = "Coral"), linewidth = 1.2) +
  geom_line(aes(y = algae, color = "Macroalgae"), linewidth = 1.2) +
  geom_point(aes(y = coral, color = "Coral"), size = 4) +
  geom_point(aes(y = algae, color = "Macroalgae"), size = 4) +
  scale_color_manual(values = c("Coral" = "#1b9e77", "Macroalgae" = "#d95f02")) +
  scale_y_continuous(limits = c(0, 50)) +
  labs(x = "Year", y = "Mean cover (%)", color = NULL,
       title = "Benthic Cover Trends") +
  theme_hri()
```

## Coral vs. Macroalgae

```{r}
full %>% filter(!is.na(tCORALavg)) %>%
  ggplot(aes(x = tFMAavg, y = tCORALavg, color = YEAR)) +
  geom_point(alpha = 0.6, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.6) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Macroalgae cover (%)", y = "Coral cover (%)", color = "Year",
       title = "Coral vs. Macroalgae") +
  theme_hri()
```

## Commercial Fish — By Family

```{r}
full %>% filter(!is.na(tCommBiomass)) %>%
  select(YEAR, tLUTJavg, tSERRavg, tCARAavg, tSPHYavg, tHAEMavg) %>%
  pivot_longer(-YEAR, names_to = "fam", values_to = "bio") %>%
  mutate(fam = recode(fam, tLUTJavg = "Snappers", tSERRavg = "Groupers",
    tCARAavg = "Jacks", tSPHYavg = "Barracuda", tHAEMavg = "Grunts")) %>%
  ggplot(aes(x = fam, y = bio, fill = YEAR)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = NULL, y = "Biomass (g/100m², log)", fill = "Year",
       title = "Commercial Fish by Family") +
  theme_hri()
```

## Herbivorous Fish — By Family

```{r}
full %>% filter(!is.na(tHerbiBiomass)) %>%
  select(YEAR, tACANavg, tSCARavg) %>%
  pivot_longer(-YEAR, names_to = "fam", values_to = "bio") %>%
  mutate(fam = recode(fam, tACANavg = "Surgeonfish", tSCARavg = "Parrotfish")) %>%
  ggplot(aes(x = fam, y = bio, fill = YEAR)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = NULL, y = "Biomass (g/100m², log)", fill = "Year",
       title = "Herbivorous Fish by Family") +
  theme_hri()
```

## Fish Biomass — Trends

```{r}
avg_c <- full %>% filter(!is.na(tCommBiomass)) %>%
  group_by(YEAR) %>% summarise(avg = mean(tCommBiomass), .groups = "drop") %>%
  mutate(group = "Commercial")
avg_h <- full %>% filter(!is.na(tHerbiBiomass)) %>%
  group_by(YEAR) %>% summarise(avg = mean(tHerbiBiomass), .groups = "drop") %>%
  mutate(group = "Herbivorous")

bind_rows(avg_c, avg_h) %>%
  ggplot(aes(x = YEAR, y = avg, group = group, color = group)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4, shape = 21, fill = "white", stroke = 1.3) +
  scale_color_manual(values = c("Commercial" = "#2C3E50", "Herbivorous" = "#E74C3C")) +
  labs(x = "Year", y = "Mean biomass (g/100m²)", color = NULL,
       title = "Fish Biomass Trends") +
  theme_hri()
```

## Response Variable Correlations

```{r}
#| fig-height: 6
resp <- input %>%
  select(tCommBiomass_avg, tHerbiBiomass_avg, tCORALavg_avg, tFMAavg_avg) %>%
  rename(`Comm. fish` = 1, `Herb. fish` = 2, Coral = 3, Macroalgae = 4) %>%
  na.omit()

corrplot(cor(resp), method = "number", type = "upper", diag = FALSE,
         number.cex = 1.2, tl.cex = 1.0, tl.col = "black",
         col = colorRampPalette(c("#E74C3C", "white", "#2C3E50"))(200))
```

## Key Ecological Relationships

:::: {.columns}
::: {.column width="50%"}
```{r}
#| fig-height: 5
ggplot(input, aes(x = tCommBiomass_avg, y = tCORALavg_avg)) +
  geom_point(alpha = 0.5, color = "#2C3E50") +
  geom_smooth(method = "lm", color = "#E74C3C", se = TRUE, alpha = 0.2) +
  labs(x = "Commercial fish (g/100m²)", y = "Coral cover (%)") +
  theme_hri()
```
:::
::: {.column width="50%"}
```{r}
#| fig-height: 5
ggplot(input, aes(x = tHerbiBiomass_avg, y = tFMAavg_avg)) +
  geom_point(alpha = 0.5, color = "#2C3E50") +
  geom_smooth(method = "lm", color = "#E74C3C", se = TRUE, alpha = 0.2) +
  labs(x = "Herbivorous fish (g/100m²)", y = "Macroalgae cover (%)") +
  theme_hri()
```
:::
::::

## Spatial Distribution — Coral Cover

```{r}
input %>% filter(!is.na(tCORALavg_avg)) %>%
  ggplot(aes(x = Longitude, y = Latitude, color = tCORALavg_avg)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_viridis_c(option = "C", name = "Coral cover (%)") +
  labs(title = "Survey Sites — Coral Cover") +
  coord_quickmap() +
  theme_hri()
```

## Spatial Distribution — Herbivorous Fish

```{r}
input %>% filter(!is.na(tHerbiBiomass_avg)) %>%
  ggplot(aes(x = Longitude, y = Latitude, color = tHerbiBiomass_avg)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_viridis_c(option = "D", name = "Biomass (g/100m²)") +
  labs(title = "Survey Sites — Herbivorous Fish") +
  coord_quickmap() +
  theme_hri()
```

## Predictor Correlations — All

```{r}
#| fig-height: 6.5
hri <- read.csv(here("data", "external", "HRI_all_variables.csv"))
resp_cols <- c("FID","Longitude_","Latitude_y","Coral_cove","Algae_cove","Commercial","Herbivorou")
preds <- hri %>% select(-any_of(resp_cols)) %>%
  mutate(across(where(is.numeric), ~ifelse(. == -9999, NA, .))) %>%
  na.omit()

corrplot(cor(preds), method = "number", type = "upper", diag = FALSE,
         number.cex = 0.55, tl.cex = 0.65, tl.col = "black")
```

## Predictor Correlations — Reduced

```{r}
#| fig-height: 6.5
exclude <- c("curv_plan", "slopeslo_1", "slope_240")
preds_r <- preds %>% select(-any_of(exclude))

corrplot(cor(preds_r), method = "number", type = "upper", diag = FALSE,
         number.cex = 0.65, tl.cex = 0.75, tl.col = "black")
```

Removed: `r paste(intersect(exclude, names(preds)), collapse = ", ")`
(|r| > 0.7)

## Predictor Distributions

```{r}
#| fig-height: 6.5
preds_r %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = val)) +
  geom_histogram(bins = 25, fill = "#2C3E50", alpha = 0.7, color = "white") +
  facet_wrap(~var, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Count", title = "Predictor Distributions") +
  theme_hri() +
  theme(axis.text.x = element_text(angle = 0, size = 8))
```

## Summary

| Metric | Value |
|--------|-------|
| Survey years | 2018, 2021, 2023 |
| Total observations | `r nrow(full)` |
| Unique sites | `r nrow(input)` |
| Mean coral cover | `r round(mean(full$tCORALavg, na.rm=TRUE), 1)`% |
| Mean macroalgae cover | `r round(mean(full$tFMAavg, na.rm=TRUE), 1)`% |
| Mean commercial biomass | `r round(mean(full$tCommBiomass, na.rm=TRUE), 0)` g/100m² |
| Mean herbivorous biomass | `r round(mean(full$tHerbiBiomass, na.rm=TRUE), 0)` g/100m² |
| Retained predictors | `r ncol(preds_r)` |
