---
title: "Correlation Matrix — Oceanographic Variables"
format:
  html:
    toc: true
  pdf:
    toc: true
---

```{r setup, message=FALSE}
library(dplyr)
library(corrplot)
library(ggplot2)
library(tidyverse)
library(MASS)
library(car)
library(here)
```

```{r paths}
Fig_output_folder   <- here("results", "figures")
Table_output_folder <- here("results", "tables")
```

## Load data

Input data with response and predictor variables. The marine predictors were
created in ArcGIS using bathymetry at 10m resolution from TNC.

The external data file `HRI_all_variables.csv` must be placed in `data/external/`.
See `data/external/README.md` for details on how to obtain it.

```{r load-data}
ext_path <- here("data", "external", "HRI_all_variables.csv")
stopifnot(
 "External data file missing: data/external/HRI_all_variables.csv\nSee data/external/README.md for instructions." =
   file.exists(ext_path)
)

HRI_variables <- read.csv(ext_path)

# Standardize column names to shorter descriptive labels
HRI_variables <- HRI_variables %>%
  rename(
    x              = Latitude,
    y              = Longitude,
    cor            = Coral_cove,
    algae          = Algae_cove,
    fish_com       = Commercial,
    fish_her       = Herbivorou,
    aspect_std_curv = aspect_s_c,
    slopeslope_240 = sloslo_240,
    slopeslope     = slopeslop,
    bathy          = bz_bathy
  )
```

## Filter to oceanographic predictors only

Some raster extraction artifacts produce -9999 for NA; replace those.

```{r filter-predictors}
colum_excluir <- c("x", "y", "cor", "algae", "fish_com", "fish_her")

data <- HRI_variables[, !names(HRI_variables) %in% colum_excluir] %>%
  mutate(across(where(is.numeric), ~ifelse(. == -9999.000, NA, .)))

print(data)
```

```{r complete-cases}
data_complete <- na.omit(data)
```

## Correlation matrix (all predictors)

```{r cor-all, fig.width=6, fig.height=6}
cor_matrix <- cor(data_complete, use = "complete.obs", method = "pearson")

png(filename = file.path(Fig_output_folder, "correlation_all_vbles.png"),
    width = 6, height = 6, units = "in", res = 300)
corrplot(cor_matrix, method = "number", type = "upper",
         diag = FALSE, number.cex = 0.7)
dev.off()

# Also display inline
corrplot(cor_matrix, method = "number", type = "upper",
         diag = FALSE, number.cex = 0.7)
```

## Remove highly correlated variables (|r| > 0.7)

Removed: `curv_pro`, `curv_plan`, `slopeslope_240`

```{r remove-correlated}
exclude_vbles <- c("curv_pro", "curv_plan", "slopeslope_240")
data.cor <- data_complete[, !names(data_complete) %in% exclude_vbles]
```

## Final correlation matrix

```{r cor-final, fig.width=6, fig.height=6}
cor_matrix.2 <- cor(data.cor, use = "complete.obs", method = "pearson")

png(filename = file.path(Fig_output_folder, "correlation_final.png"),
    width = 6, height = 6, units = "in", res = 300)
corrplot(cor_matrix.2, method = "number", type = "upper",
         diag = FALSE, number.cex = 0.7)
dev.off()

corrplot(cor_matrix.2, method = "number", type = "upper",
         diag = FALSE, number.cex = 0.7)
```

## Export model-ready data

```{r export}
data_model <- HRI_variables[, !names(HRI_variables) %in% exclude_vbles]

write.csv(data_model,
          file = file.path(Table_output_folder, "bz_data_model.csv"),
          row.names = FALSE)
```
