---
title: "HRI AGRRA Belize — Species-Level EDA"
subtitle: "Individual fish family biomass distributions, trends, and spatial patterns"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Code"
    embed-resources: true
    fig-width: 10
    fig-height: 7
  pdf:
    documentclass: article
    papersize: letter
    margin-left: 1in
    margin-right: 1in
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
  echo: true
editor: visual
---

```{r setup}
#| echo: false
library(tidyverse)
library(here)
library(corrplot)

fig_dir <- here("results", "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

theme_hri <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 16),
      plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray50"),
      axis.title    = element_text(face = "bold"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_line(color = "gray85", linetype = "dashed"),
      panel.grid.minor = element_blank(),
      legend.position  = "top",
      legend.text      = element_text(size = 11)
    )
}

# --- Load data ----------------------------------------------------------------
full  <- read.csv(here("results", "tables", "ResponseVariables_full.csv"))
input <- read.csv(here("results", "tables", "ResponseVariables_input.csv"))
full$YEAR <- factor(full$YEAR)

# Species lookup table
species_info <- tribble(
  ~code,       ~family,          ~common,        ~group,
  "tLUTJavg",  "Lutjanidae",     "Snappers",     "Commercial",

  "tSERRavg",  "Serranidae",     "Groupers",     "Commercial",
  "tCARAavg",  "Carangidae",     "Jacks",        "Commercial",
  "tSPHYavg",  "Sphyraenidae",   "Barracuda",    "Commercial",
  "tHAEMavg",  "Haemulidae",     "Grunts",       "Commercial",
  "tACANavg",  "Acanthuridae",   "Surgeonfish",  "Herbivorous",
  "tSCARavg",  "Scaridae",       "Parrotfish",   "Herbivorous"
)

species_cols <- species_info$code
species_labels <- setNames(species_info$common, species_info$code)
group_colors <- c("Commercial" = "#2C3E50", "Herbivorous" = "#E74C3C")
```

## Overview

This document examines individual fish family biomass patterns across the Belize Barrier Reef, moving beyond the grouped commercial/herbivorous totals used in the initial analysis. Seven fish families are examined: 5 commercial (Snappers, Groupers, Jacks, Barracuda, Grunts) and 2 herbivorous (Surgeonfish, Parrotfish).

```{r data-summary}
# Observations with fish data
fish_data <- full %>% filter(!is.na(tCommBiomass) | !is.na(tHerbiBiomass))

fish_data %>%
  summarise(
    `Total obs` = n(),
    `With commercial` = sum(!is.na(tCommBiomass)),
    `With herbivorous` = sum(!is.na(tHerbiBiomass)),
    `Years` = paste(unique(YEAR), collapse = ", ")
  ) %>%
  knitr::kable()
```

## Species Distributions

### Biomass distributions by family

```{r species-distributions}
#| fig-height: 8
fish_long <- full %>%
  select(YEAR, all_of(species_cols)) %>%
  pivot_longer(-YEAR, names_to = "species", values_to = "biomass") %>%
  filter(!is.na(biomass)) %>%
  mutate(
    common = species_labels[species],
    group  = species_info$group[match(species, species_info$code)]
  )

ggplot(fish_long, aes(x = biomass + 1, fill = group)) +
  geom_histogram(bins = 40, alpha = 0.7, color = "white") +
  scale_x_log10(labels = scales::comma) +
  scale_fill_manual(values = group_colors) +
  facet_wrap(~common, scales = "free", ncol = 2) +
  labs(x = "Biomass (g/100m², log scale)", y = "Count", fill = "Group",
       title = "Biomass Distributions by Fish Family") +
  theme_hri()
```

### Summary statistics

```{r species-summary-table}
fish_long %>%
  group_by(common, group) %>%
  summarise(
    n       = n(),
    mean    = round(mean(biomass, na.rm = TRUE), 1),
    median  = round(median(biomass, na.rm = TRUE), 1),
    sd      = round(sd(biomass, na.rm = TRUE), 1),
    min     = round(min(biomass, na.rm = TRUE), 1),
    max     = round(max(biomass, na.rm = TRUE), 1),
    pct_zero = round(100 * mean(biomass == 0, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(group, desc(mean)) %>%
  knitr::kable(col.names = c("Family", "Group", "n", "Mean", "Median", "SD",
                              "Min", "Max", "% Zero"))
```

### Boxplots by family and year

```{r species-boxplots-year}
#| fig-height: 8
ggplot(fish_long, aes(x = common, y = biomass + 1, fill = YEAR)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~group, scales = "free_x") +
  labs(x = NULL, y = "Biomass (g/100m², log scale)", fill = "Year",
       title = "Fish Biomass by Family and Year") +
  theme_hri()
```

## Temporal Trends

### Mean biomass by family over time

```{r species-trends}
#| fig-height: 6
trends <- fish_long %>%
  group_by(YEAR, common, group) %>%
  summarise(mean_bio = mean(biomass, na.rm = TRUE), .groups = "drop")

ggplot(trends, aes(x = YEAR, y = mean_bio, group = common, color = common)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3, shape = 21, fill = "white", stroke = 1.2) +
  facet_wrap(~group, scales = "free_y") +
  scale_color_brewer(palette = "Set2") +
  labs(x = "Year", y = "Mean biomass (g/100m²)", color = "Family",
       title = "Biomass Trends by Fish Family") +
  theme_hri()
```

### Relative composition over time

```{r species-composition}
comp <- fish_long %>%
  group_by(YEAR, group, common) %>%
  summarise(total = sum(biomass, na.rm = TRUE), .groups = "drop") %>%
  group_by(YEAR, group) %>%
  mutate(pct = 100 * total / sum(total))

ggplot(comp, aes(x = YEAR, y = pct, fill = common)) +
  geom_col(alpha = 0.85, color = "white", linewidth = 0.3) +
  facet_wrap(~group) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Year", y = "% of Group Biomass", fill = "Family",
       title = "Relative Composition within Groups") +
  theme_hri()
```

## Species Correlations

### Pairwise correlations among all 7 families

```{r species-correlation-matrix}
#| fig-height: 7
fish_wide <- full %>%
  select(all_of(species_cols)) %>%
  rename_with(~species_labels[.x]) %>%
  na.omit()

cor_mat <- cor(fish_wide, use = "pairwise.complete.obs")

corrplot(cor_mat, method = "number", type = "upper", diag = FALSE,
         number.cex = 0.85, tl.cex = 0.9, tl.col = "black",
         col = colorRampPalette(c("#E74C3C", "white", "#2C3E50"))(200),
         title = "Pairwise Correlations — Fish Family Biomass",
         mar = c(0, 0, 2, 0))
```

### Correlations with benthic cover

```{r species-benthic-correlations}
#| fig-height: 6
benthic_fish <- full %>%
  select(tCORALavg, tFMAavg, all_of(species_cols)) %>%
  rename(Coral = tCORALavg, Macroalgae = tFMAavg) %>%
  rename_with(~ifelse(.x %in% names(species_labels), species_labels[.x], .x)) %>%
  na.omit()

cor_bf <- cor(benthic_fish)

corrplot(cor_bf, method = "number", type = "upper", diag = FALSE,
         number.cex = 0.7, tl.cex = 0.85, tl.col = "black",
         col = colorRampPalette(c("#E74C3C", "white", "#2C3E50"))(200),
         title = "Correlations: Fish Families × Benthic Cover",
         mar = c(0, 0, 2, 0))
```

### Key scatterplots

```{r species-scatter-key}
#| fig-height: 8
p_pairs <- list()

# Parrotfish vs macroalgae (herbivory hypothesis)
p_pairs[[1]] <- ggplot(full, aes(x = tSCARavg, y = tFMAavg, color = YEAR)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Parrotfish biomass", y = "Macroalgae cover (%)",
       title = "Parrotfish vs Macroalgae") +
  theme_hri()

# Groupers vs snappers (competition/predation)
p_pairs[[2]] <- ggplot(full, aes(x = tSERRavg, y = tLUTJavg, color = YEAR)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Grouper biomass", y = "Snapper biomass",
       title = "Groupers vs Snappers") +
  theme_hri()

# Parrotfish vs coral (bioerosion/facilitation)
p_pairs[[3]] <- ggplot(full, aes(x = tSCARavg, y = tCORALavg, color = YEAR)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Parrotfish biomass", y = "Coral cover (%)",
       title = "Parrotfish vs Coral") +
  theme_hri()

# Surgeonfish vs parrotfish (herbivore competition)
p_pairs[[4]] <- ggplot(full, aes(x = tACANavg, y = tSCARavg, color = YEAR)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Surgeonfish biomass", y = "Parrotfish biomass",
       title = "Surgeonfish vs Parrotfish") +
  theme_hri()

gridExtra::grid.arrange(grobs = p_pairs, ncol = 2)
```

## Spatial Patterns

### Individual species spatial distributions

```{r species-spatial-maps}
#| fig-height: 10
# Site-level averages for mapping
site_avg <- full %>%
  group_by(Name, Latitude, Longitude) %>%
  summarise(across(all_of(species_cols), ~mean(.x, na.rm = TRUE)),
            .groups = "drop")

site_long <- site_avg %>%
  pivot_longer(all_of(species_cols), names_to = "species", values_to = "biomass") %>%
  mutate(common = species_labels[species],
         group  = species_info$group[match(species, species_info$code)])

ggplot(site_long, aes(x = Longitude, y = Latitude, color = biomass + 1)) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_viridis_c(option = "C", trans = "log10", name = "Biomass\n(g/100m²)") +
  facet_wrap(~common, ncol = 2) +
  coord_quickmap() +
  labs(title = "Spatial Distribution — Fish Family Biomass") +
  theme_hri() +
  theme(axis.text.x = element_text(angle = 0, size = 8))
```

### Latitude gradients by species

```{r species-latitude-gradient}
#| fig-height: 8
site_long %>%
  mutate(lat_band = cut(Latitude, breaks = 5, labels = c("South", "S-Mid",
                                                           "Central", "N-Mid", "North"))) %>%
  ggplot(aes(x = lat_band, y = biomass + 1, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_manual(values = group_colors) +
  facet_wrap(~common, scales = "free_y", ncol = 2) +
  labs(x = "Latitude Band", y = "Biomass (g/100m², log)", fill = "Group",
       title = "Latitude Gradients by Species") +
  theme_hri()
```

## Dominance and Evenness

### Which families dominate at each site?

```{r dominance-analysis}
#| fig-height: 6
dominant <- site_avg %>%
  pivot_longer(all_of(species_cols), names_to = "species", values_to = "biomass") %>%
  group_by(Name, Latitude, Longitude) %>%
  slice_max(biomass, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(common = species_labels[species])

ggplot(dominant, aes(x = Longitude, y = Latitude, color = common)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_brewer(palette = "Set2") +
  coord_quickmap() +
  labs(color = "Dominant Family",
       title = "Dominant Fish Family by Site") +
  theme_hri()
```

```{r dominance-table}
dominant %>%
  count(common) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n)) %>%
  knitr::kable(col.names = c("Dominant Family", "Sites", "% of Sites"))
```

### Simpson's evenness across families

```{r evenness}
#| fig-height: 5
evenness <- site_avg %>%
  rowwise() %>%
  mutate(
    total = sum(c_across(all_of(species_cols)), na.rm = TRUE),
    simpson_d = 1 - sum((c_across(all_of(species_cols)) / total)^2, na.rm = TRUE),
    evenness = simpson_d / (1 - 1/7)  # normalized by max possible
  ) %>%
  ungroup()

ggplot(evenness, aes(x = Longitude, y = Latitude, color = evenness)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "D", name = "Evenness") +
  coord_quickmap() +
  labs(title = "Fish Community Evenness (Simpson's)") +
  theme_hri()
```

## Predictor Relationships

### Species biomass vs bathymetric predictors

```{r species-vs-predictors}
#| fig-height: 10
hri <- read.csv(here("data", "external", "HRI_all_variables.csv"))

# Merge individual species with predictors on coordinates
site_pred <- site_avg %>%
  inner_join(
    hri %>% rename(Longitude = Longitude_, Latitude = Latitude_y),
    by = c("Longitude", "Latitude")
  )

# Depth relationships
depth_long <- site_pred %>%
  select(depth, all_of(species_cols)) %>%
  pivot_longer(-depth, names_to = "species", values_to = "biomass") %>%
  mutate(common = species_labels[species])

ggplot(depth_long, aes(x = depth, y = biomass + 1)) +
  geom_point(alpha = 0.3, size = 1, color = "#2C3E50") +
  geom_smooth(method = "loess", color = "#E74C3C", se = TRUE, alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~common, scales = "free_y", ncol = 2) +
  labs(x = "Depth (m)", y = "Biomass (g/100m², log scale)",
       title = "Fish Family Biomass vs Depth") +
  theme_hri() +
  theme(axis.text.x = element_text(angle = 0))
```

### Slope relationships

```{r species-vs-slope}
#| fig-height: 10
slope_long <- site_pred %>%
  select(slope, all_of(species_cols)) %>%
  pivot_longer(-slope, names_to = "species", values_to = "biomass") %>%
  mutate(common = species_labels[species])

ggplot(slope_long, aes(x = slope, y = biomass + 1)) +
  geom_point(alpha = 0.3, size = 1, color = "#2C3E50") +
  geom_smooth(method = "loess", color = "#E74C3C", se = TRUE, alpha = 0.2) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~common, scales = "free_y", ncol = 2) +
  labs(x = "Terrain Slope", y = "Biomass (g/100m², log scale)",
       title = "Fish Family Biomass vs Slope") +
  theme_hri() +
  theme(axis.text.x = element_text(angle = 0))
```

## Summary

```{r summary-table}
fish_long %>%
  group_by(common, group) %>%
  summarise(
    `Mean biomass` = round(mean(biomass, na.rm = TRUE), 1),
    `Dominant at n sites` = sum(dominant$common == first(common)),
    .groups = "drop"
  ) %>%
  arrange(group, desc(`Mean biomass`)) %>%
  knitr::kable()
```

**Key observations:**

-   Individual species show distinct spatial patterns not captured by grouped totals
-   Parrotfish and Grunts tend to dominate biomass at most sites
-   Species-specific depth preferences suggest niche partitioning along the depth gradient
-   Pairwise correlations among families reveal both positive (co-occurrence) and negative (competitive exclusion) relationships
-   Community evenness varies spatially, with potential north-south gradients along the barrier reef
